---
title: "ST3189 Coursework"
author: "Andrea Kojovic"
date: "2025-02-04"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \setmainfont{Arial}  # Use a font that supports Unicode
geometry: margin=1in
always_allow_html: true

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r, warning=FALSE}

# Install packages
package <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)}


# Write the name of your packages here
packages <- c("dplyr","ggplot2","readr","MASS","ggcorrplot","clustMixType","factoextra","plotly","cluster","clValid","mclust","reshape2","tidyr","dendextend","glmnet","GGally","ggpubr","gridExtra","car","randomForest","pdp","ROSE","e1071","pROC","PRROC","caret","rlang")

package(packages)

```

```{r, results='hide'}
#Call the necessary libraries before proceeding
lapply(packages, library, character.only = TRUE)

``` 

# Load the dataset
```{r}

# Load the dataset
df <- read.csv("C:/Users/USER/Downloads/archive/BankChurners.csv")

# View the first few rows
head(df)

# Check the structure of the dataset
str(df)

# Get column names
colnames(df)

# Summary statistics
summary(df)

```

#Remove unnecessary columns

Removal of unnecessary columns in the dataset that are not relevant for analysis.

```{r}
# Drop unnecessary columns
df <- df %>%
  select(-c((ncol(df)-1):ncol(df)))


# Verify the dataset after removal
str(df)

```
Checking for any missing values in the dataset. 

```{r}
# Count missing values in each column
colSums(is.na(df))

```
#Imputation of the 'Unknown' values   

Since some of the categorical variables like 'Marital Status', 'Education Level' and others have 'Unknown' values, we use KNN to impute these values and avoid removing these data points altogether. This type of financial data tends to follow similar pattern, hence the imputation.  

```{r}
# Identify categorical columns
cat_cols <- names(df)[sapply(df, is.character) | sapply(df, is.factor)]

# Identify which categorical columns contain 'Unknown'
unknown_cols <- sapply(df[cat_cols], function(x) any(x == "Unknown"))
unknown_cols <- names(unknown_cols[unknown_cols == TRUE])

# Replace 'Unknown' values with NA only in the affected columns
df[unknown_cols] <- lapply(df[unknown_cols], function(x) {
  x <- as.character(x)  # Convert factor to character
  x[x == "Unknown"] <- NA  # Replace only 'Unknown' with NA
  return(as.factor(x))  # Convert back to factor
})

# Check missing values
colSums(is.na(df[unknown_cols]))

```

```{r}
# Store original factor levels before conversion
cat_levels <- lapply(df[unknown_cols], levels)

# Convert only affected categorical columns to numeric
df_encoded <- df
for (col in unknown_cols) {
  df_encoded[[col]] <- as.numeric(factor(df[[col]], levels=cat_levels[[col]]))  # Convert only columns with missing values
}

# Verify encoding
str(df_encoded[unknown_cols])

```

```{r}

# Create a backup of original values
df_backup <- df_encoded

# Apply KNN Imputation using all available columns
preproc <- preProcess(df_encoded, method="knnImpute", k=5)
df_imputed <- predict(preproc, newdata=df_encoded)

# Restore original non-missing values
for (col in unknown_cols) {
  df_imputed[[col]][!is.na(df_backup[[col]])] <- df_backup[[col]][!is.na(df_backup[[col]])]
}

# Verify that only missing values were changed
colSums(is.na(df_imputed))  # Should return 0 for all columns

```
```{r}
df_final <- df  # Copy original dataset

# Convert imputed numeric values back to categorical
for (col in unknown_cols) {
  df_imputed[[col]] <- round(df_imputed[[col]])  # Round to nearest integer
  df_imputed[[col]] <- pmin(pmax(df_imputed[[col]], 1), length(cat_levels[[col]]))  # Ensure within valid range
  df_final[[col]] <- factor(df_imputed[[col]], levels = 1:length(cat_levels[[col]]), labels = cat_levels[[col]])
}

# Verify structure after restoring categorical values
str(df_final[unknown_cols])

```

```{r}
for (col in unknown_cols) {
  print(table(df_final[[col]], useNA="ifany"))
}

```


#Exploratory data analysis 

We visualise the variables and their distribution to check for skewness, variance and outliers. 

```{r}
num_cols <- names(df_final)[sapply(df_final, is.numeric)]

# Save histograms
for (col in num_cols) {
  png(filename = paste0("plots/hist_", col, ".png"), width = 800, height = 600)
  hist(df_final[[col]], main = col, xlab = col, col = "lightblue", border = "black", breaks = 30)
  file_name <- paste0("histogram_", col, ".png")
  dev.off()  # Close file
}

# Save boxplots
for (col in num_cols) {
  p <- ggplot(df_final, aes(x = "", y = !!sym(col))) +
    geom_boxplot(fill = "lightblue", outlier.color = "red") +
    labs(title = paste("Boxplot of", col), y = col) +
    theme_minimal()
  
  # Save the plot to a file
  file_name <- paste0("boxplot_", col, ".png")
  ggsave(file_name, plot = p, width = 6, height = 4)
}
```

```{r}

# Visualize distributions and check for scaling needs
par(mfrow=c(3,3))  # Adjust the grid layout as needed
for (col in num_cols) {
  hist(df_final[[col]], main=col, xlab=col, col="lightblue", border="black", breaks=30)
}
par(mfrow=c(1,1))  # Reset plot layout

# Boxplots to detect outliers visually
for (col in num_cols) {
  print(ggplot(df_final, aes(x = "", y = !!sym(col))) +
          geom_boxplot(fill = "lightblue", outlier.color = "red") +
          labs(title = paste("Boxplot of", col), y = col) +
          theme_minimal())
}

```

# Outlier detection and removal for skewed variables 

It is noticeable that some of the variables in the dataset are skewed and contain significant outliers. We use a simple IQR method for outlier detection and removal. 

```{r}

skewed_vars <- c("Avg_Open_To_Buy", "Total_Trans_Amt", "Total_Trans_Ct", "Avg_Utilization_Ratio","Credit_Limit","Total_Revolving_Bal")

for (var in skewed_vars) {
    Q1 <- quantile(df_final[[var]], 0.25)
    Q3 <- quantile(df_final[[var]], 0.75)
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    
    # Remove outliers outside the bounds
    df_final <- df_final[df_final[[var]] >= lower_bound & df_final[[var]] <= upper_bound, ]
}
```

## Visualisation after outlier removal 

We now visualise the variables after the outlier removal to assess the changes and whether any significant distribution changes have been done. 

```{r}

# Visualizing transformed variables using histograms
par(mfrow=c(3,3))  # Arrange multiple plots in a grid

for (col in skewed_vars) {
  hist(df_final[[col]], main=paste("Transformed", col), xlab=col, col="lightblue", border="black", breaks=30)
}

par(mfrow=c(1,1))  # Reset plot layout

```

```{r}

for (col in skewed_vars) {
  png(filename = paste0("plots/hist_", col, ".png"), width = 800, height = 600)
  hist(df_final[[col]], main = paste("Cleaned", col), xlab = col, col = "lightblue", border = "black", breaks = 30)
  file_name <- paste0("cleaned_histogram_", col, ".png")
  dev.off()  # Close file
}

```

#Variable transformation and scaling

To make the data appropriate for analysis we standardise the normally distributed variables to a standard normal distribution and perform a log transformation for the skewed variables. 

```{r}

# Variables to standardize (normally distributed)
standardize_vars <- c("Customer_Age", "Months_on_book", "Total_Trans_Ct", "Total_Ct_Chng_Q4_Q1", "Total_Amt_Chng_Q4_Q1", "Dependent_count","Total_Relationship_Count","Months_Inactive_12_mon","Contacts_Count_12_mon")

# Variables to log-transform (right-skewed)
log_transform_vars <- c("Credit_Limit", "Avg_Open_To_Buy", "Total_Revolving_Bal", "Total_Trans_Amt", "Avg_Utilization_Ratio")

# Copy dataset
df_transformed <- df_final

# Step 1: Log transformation (only for skewed variables)
df_transformed[log_transform_vars] <- log(df_transformed[log_transform_vars] + 1) # Adding 1 to avoid log(0) issues

# Step 2: Standardization (Z-score scaling for both normal variables)
df_transformed[standardize_vars] <- scale(df_transformed[standardize_vars])

# Check summary statistics
summary(df_transformed)


```

##Visualise after scaling

After transformations it is necessary to visualise the variables againa and assess how appropriate they are for analysis, but also to check for inconsistencies and mistakes in transformation. 

```{r}
# Visualizing transformed variables using histograms
par(mfrow=c(3,3))  # Arrange multiple plots in a grid

for (col in num_cols) {
  hist(df_transformed[[col]], main=paste("Transformed", col), xlab=col, col="lightblue", border="black", breaks=30)
}

par(mfrow=c(1,1))  # Reset plot layout

```

```{r}

for (col in num_cols) {
  png(filename = paste0("plots/hist_", col, ".png"), width = 800, height = 600)
  hist(df_transformed[[col]], main=paste("Transformed", col), xlab=col, col="lightblue", border="black", breaks=30)
  file_name <- paste0("transformed_histogram_", col, ".png")
  dev.off()  # Close file
}


```


#Outlier detection and removal

Since there are notable outliers detected in the visualisation of the numerical variables, it is necessary to remove these in order to improve the analysis. 

```{r}


# Identify rows with extreme Z-scores (greater than 3 or less than -3)
outlier_rows <- which(abs(df_transformed[standardize_vars]) > 3, arr.ind = TRUE)

# Remove outliers
clean_data <- df_transformed[-outlier_rows, ]

```

## Visualise after outlier removal 

Finally, we visualise all the variables after transformation, scaling and outlier detection. 

```{r}
# Visualizing transformed variables using histograms
par(mfrow=c(3,3))  # Arrange multiple plots in a grid

for (col in num_cols) {
  hist(clean_data[[col]], main=paste("Transformed", col), xlab=col, col="lightblue", border="black", breaks=30)
}

par(mfrow=c(1,1))  # Reset plot layout

```

```{r}

for (col in num_cols) {
  png(filename = paste0("plots/hist_", col, ".png"), width = 800, height = 600)
  hist(clean_data[[col]], main=paste("Transformed and Cleaned", col), xlab=col, col="lightblue", border="black", breaks=30)
  file_name <- paste0("transformed_and_cleaned_histogram_", col, ".png")
  dev.off()  # Close file


}

```

#Checking for correlation

Before performing regression and classification, it is necessary to check for correlations between variables to remove any highly correlated variables and do feature engineering if appropriate. 

```{r}

# Compute the correlation matrix for numerical variables
cor_matrix <- cor(df_final[num_cols], use = "complete.obs")  # Compute correlation using complete observations

# Print the correlation matrix
print(cor_matrix)

# Visualize the correlation matrix using ggcorrplot
corr_plot <- ggcorrplot(cor_matrix, 
                         method = "square", 
                         type = "lower", 
                         lab = TRUE, 
                         lab_size = 3, 
                         colors = c("blue", "white", "red"),
                         title = "Correlation Matrix of Numerical Variables",
                         ggtheme = theme_minimal())

# Display the plot (optional)
print(corr_plot)

# Save the plot as a PNG file
ggsave("correlation_plot.png", plot = corr_plot, width = 6, height = 5)

```

Instead of the transformed data correlation, the raw data correlation matrix was computed and the only significant correlations we see are:
-Avg open to buy and credit limit - The "Avg Open to Buy" could be correlated with the credit limit because a customer's credit limit is what defines their available spending power. If a customer has a high credit limit, they may have a higher Avg Open to Buy because they have more credit available to them. Similarly, if their credit limit is lower or they have used a significant portion of their available credit, their Avg Open to Buy would be lower.
-Total transaction amount and Total transaction count - makes sense that the more transactions a customer makes, the higher the transaction amount 
-Months on book and Customer Age - makes sense that older customers have had their accounts for longer and vice versa
-Total revolving balance and Avg Utilization ratio - A higher ratio (generally over 30-40%) could indicate that the customer is close to their credit limit

#Clustering

We begin with clustering algorithms to first understand any significant customer groups relevant to the bank. It helps identify the key features and the characteristics of these groups which may aid further analysis of regression and classification. 

##Preprocessing 

It is necessary to preprocess the clustering dataset before performing analysis to exclude irrelevant columns like client number and gender. It is also relevant to transform categorical variables to factors and ensure the numerical variables are identified as numeric. 

```{r}

# Exclude irrelevant columns
df_clustering <- clean_data[, -c(1, 2)]  # Excludes first two columns
df_clustering <- df_clustering %>% dplyr::select(-Gender)  # Then exclude 'Gender'
  # Exclude the first two columns and 'Gender'

# Identify categorical variables
cat_cols <- names(df_clustering)[sapply(df_clustering, is.character)]
df_clustering[cat_cols] <- lapply(df_clustering[cat_cols], as.factor)  # Convert to factors

# Identify numerical variables
num_cols <- names(df_clustering)[sapply(df_clustering, is.numeric)]
df_clustering[num_cols] <- scale(df_clustering[num_cols])  # Standardize numerical variables

# Print structure
str(df_clustering)


```

##Optimal number of clusters

Before proceeding with k-means clustering, we visualise the elbow plot and calculate silouette scores for different number of clusters to find the optimal. 

```{r}

# Test different k values (adjust max to 10)
wss <- numeric(10)  # Within-cluster sum of squares
for (k in 2:10) {
  wss[k] <- sum(kproto(df_clustering, k)$tot.withinss)
}


# Plot the Elbow Method
plot(2:10, wss[2:10], type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K", ylab = "Total Within Sum of Squares",
     main = "Elbow Method for Optimal Clusters")

# Open a PNG device, create and save the Elbow plot, then close the device
png("elbow_plot.png", width = 800, height = 600)
plot(2:10, wss[2:10], type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K", ylab = "Total Within Sum of Squares",
     main = "Elbow Method for Optimal Clusters")
dev.off()

# Silhouette Score (More Reliable)
silhouette_scores <- numeric(10)
for (k in 2:10) {
  silhouette_scores[k] <- mean(silhouette(kproto(df_clustering, k)$cluster, dist(df_clustering))[, 3])
}

plot(2:10, silhouette_scores[2:10], type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K", ylab = "Average Silhouette Score",
     main = "Silhouette Score for Optimal Clusters")

# Save the Silhouette Plot as a PNG file
png("silhouette_plot.png", width = 800, height = 600)
plot(2:10, silhouette_scores[2:10], type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K", ylab = "Average Silhouette Score",
     main = "Silhouette Score for Optimal Clusters")
dev.off()

```
In both the elbow plot, the curve starts decreasing less slowly and is identified as the 'elbow' of the plot and therefore the optimal number of clusters. Similarly, the silouette scores show a significant drop before 4, and the gradient significantly decreases thereafter. Therefore, the optimal number of clusters is 4, which is used for the k-means clustering. 

##Apply K-Prototypes clustering on mixed data

K-Prototypes clustering is performed on mixed data - both numeric and categorical variables since the categorical variables are relevant for the analysis. 

```{r}
set.seed(42)  # For reproducibility

optimal_k <- 4  # Adjust based on the elbow/silhouette method

# Run K-Prototypes clustering
kproto_model <- kproto(df_clustering, optimal_k)

# Add cluster labels
df_clustering$Cluster <- as.factor(kproto_model$cluster)

# Print cluster sizes
table(df_clustering$Cluster)

```
4 clusters are identified, with a relatively similar number of data points, aside from cluster 4. 

##Cluster visualisation

The key of the clustering is to attempt to visualise the different clusters and their differences, as well as how 'far apart' they are from each other. 

```{r}

# Reduce dimensions using PCA
pca <- prcomp(df_clustering[num_cols], center = TRUE, scale. = TRUE)
pca_df <- as.data.frame(pca$x)
pca_df$Cluster <- df_clustering$Cluster

# Scatter plot of clusters in PCA space
cluster_scatter_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Customer Segments (PCA Projection)", x = "PC1", y = "PC2")

print(cluster_scatter_plot)

ggsave("pca_scatter_plot.png", plot = cluster_scatter_plot, width = 8, height = 6, dpi = 300)

```
PCA is performed in order to identify principal components for the purpose of visualisation. The clusters do not seem to be far apart in 2D. However, the clusters are clearly grouped together, with the exception of cluster 1 which not only seems to have little data points visually, but also seems to coincide with cluster 3. 

```{r}

# Create a 3D plot using plotly
pca_3d_plot <- plot_ly(pca_df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~Cluster, 
                       colors = c('#1f77b4', '#ff7f0e', '#2ca02c', '#d62728', '#9467bd', '#8c564b'),
                       type = 'scatter3d', mode = 'markers', marker = list(size = 5)) %>%
  layout(title = "Customer Segments (3D PCA Projection)",
         scene = list(xaxis = list(title = "PC1"),
                      yaxis = list(title = "PC2"),
                      zaxis = list(title = "PC3")))

# Show the plot
pca_3d_plot

```

The 3D makes the distinction between the clusters easier to see. The second component here helps distinguish between clusters 1,3 and 4. 

##Analysing cluster characteristics

It is important to analyse the characteristics of each cluster, that is the average or most common (categorical variables) values for each of the cluster and how this differentiates them. This helps identify 

```{r}
num_summary <- df_clustering %>%
  group_by(Cluster) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

print(num_summary)

mode_func <- function(x) {as.character(names(sort(table(x), decreasing = TRUE)[1]))}

cat_summary <- df_clustering %>%
  group_by(Cluster) %>%
  summarise(across(where(is.factor), mode_func))

print(cat_summary)

```

In order to separate the clusters by some differentiating factors, we analyse the average values across variables in each of the clusters. Looking at customer age, clusters 1 and 2 seem to have the same average age - which is relatively low, cluster 3 right above and cluster 4 significantly higher than the rest. Looking at dependent count, clusters 1,2 and 3 have slightly above average, hence likely being adults with children. Cluster 4 has a dependent count lower than average, thus likely being older individuals without children or with grown children. Months on book and total relationship count variables provide similar insights as customer age. From months inactive in the last 12 months, the values seem to be fairly similar, aside from cluster 1 being higher than others. Clusters 1 and 4 have more contact with the bank than clusters 2 and 3 (with 3 being the lowest). Credit limits differ significantly across the clusters, with cluster 2 having the lowest, followed by cluster 1. Conversely, clusters 3 and 4 have significantly higher than others, with cluster 3 being highest. The total revolving balance, however, is highest in cluster 2, followed by clusters 4,3 and the lowest being 1.  Consequently, a similar pattern exists in the average utilization ratio. Finally, clusters 2 and 3 both have high (and similar) transaction count and amount. Cluster 1 follows with significantly lower values and finally cluster 4 with significantly low amounts. 

Cluster 1: "Young Adults with High Engagement and Moderate Finances"

Characteristics:

Age: Relatively young, similar to Cluster 2, likely adults in the early stages of their careers.

Dependent Count: Slightly above average, indicating that they may be starting families or have young children.

Months on Book & Total Relationship Count: Both values are moderate, suggesting these individuals are still establishing their relationship with the bank.

Months Inactive in the Last 12 Months: Slightly higher than other clusters, indicating they might be more sporadic in their engagement.

Credit Limit: Low, possibly due to lower credit scores or less established financial profiles.

Total Revolving Balance: Higher than others, indicating that while they have lower credit limits, they may be using a larger proportion of their available credit.

Utilization Ratio: High, aligning with the high revolving balance, indicating they rely more on credit for their purchases.

Transaction Count & Amount: Low, which suggests limited spending or perhaps a focus on saving or budgeting.

Cluster 2: "Young Families with Moderate Finances and High Transactional Activity"

Characteristics:

Age: Relatively young, likely in their 30s to early 40s, likely to be balancing family and work life.

Dependent Count: Slightly above average, indicating families with children or dependents.

Months on Book & Total Relationship Count: Moderate values, suggesting they are not new to the bank but are still in the process of developing their full financial relationship.

Months Inactive in the Last 12 Months: Similar to other clusters, indicating relatively stable engagement with the bank.

Credit Limit: Low credit limit compared to others, likely reflecting financial responsibility in the early family years.

Total Revolving Balance: Higher than others, reflecting moderate debt use, likely tied to family-related expenses.

Utilization Ratio: High, which is a concern given the higher balance.

Transaction Count & Amount: High transactional activity, showing that these customers have frequent interactions with their bank, possibly driven by family-related purchases.

Cluster 3: "Established Professionals with High Credit Use and Transaction Activity"

Characteristics:

Age: Slightly higher than Clusters 1 and 2, indicating established professionals in their late 30s to 40s.

Dependent Count: Slightly above average, indicating families but with older children.

Months on Book & Total Relationship Count: Higher values, showing that these individuals have long-term relationships with the bank.

Months Inactive in the Last 12 Months: Similar to the other clusters, implying moderate engagement.

Credit Limit: Highest in the dataset, indicating a strong financial profile or high earning potential.

Total Revolving Balance: Moderate, but their high credit limits imply a reliance on credit.

Utilization Ratio: High, indicating that despite their high credit limits, they are frequently using a large portion of it.

Transaction Count & Amount: Both values are significantly high, indicating a high level of activity, likely from regular spending or business transactions.

Cluster 4: "Older, Less Engaged Customers with Low Financial Activity"

Characteristics:

Age: Significantly higher than other clusters, likely customers in their 50s to 60s or older.

Dependent Count: Below average, possibly reflecting empty nesters or individuals without dependents.

Months on Book & Total Relationship Count: Both high, indicating long-standing relationships with the bank.

Months Inactive in the Last 12 Months: Similar to other clusters, suggesting some level of disengagement or stable, passive account management.

Credit Limit: Relatively high, possibly due to a long banking history.

Total Revolving Balance: Lower compared to others, showing less reliance on credit.

Utilization Ratio: Low, which aligns with their lower balance and less reliance on credit.

Transaction Count & Amount: Low, indicating that they are not actively spending.

##Further inspection of clusters 

```{r}

# Compute silhouette score per observation
sil <- silhouette(kproto_model$cluster, dist(df_clustering))

# Visualize silhouette plot
sil_scores <- fviz_silhouette(sil) + ggtitle("Silhouette Plot of Clusters")

# Display the plot in your R Markdown document
print(sil_scores)

# Save the silhouette plot as a PNG file
ggsave("silhouette_scores.png", plot = sil_scores, width = 8, height = 6, dpi = 300) 

# Compute Dunn Index
dunn_score <- dunn(dist(df_clustering), kproto_model$cluster)
print(dunn_score)

# Compare clustering runs
set.seed(123)
kproto_model2 <- kproto(df_clustering, 4)  # Run again
ari_score <- adjustedRandIndex(kproto_model$cluster, kproto_model2$cluster)

print(ari_score)

```

```{r}

# Melt dataset for ggplot (long format)
df_long <- melt(df_clustering, id.vars = "Cluster")

# Plot density distribution for numeric variables
density_plot <- ggplot(df_long, aes(x = value, fill = Cluster)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Variable Distributions by Cluster", x = "Value", y = "Density")

print(density_plot)

ggsave("density_distribution_by_cluster.png", plot = density_plot, width = 10, height = 8, dpi = 300)

```

It is also useful to visualise the categorical variables across the 4 clusters to gain deeper insights into their characteristics. 

```{r}
# Plot categorical distributions
cat_cols <- names(df_clustering)[sapply(df_clustering, is.factor)]

for (col in cat_cols) {
  # Create the plot for the current categorical variable
  categorical_plot<- ggplot(df_clustering, aes_string(x = col, fill = "Cluster")) +
         geom_bar(position = "fill") +
         theme_minimal() +
         labs(title = paste("Cluster Proportions for", col),
              x = col, y = "Proportion")
  
  # Print the plot to display in the R Markdown document
  print(categorical_plot)
  
  # Save the plot as a PNG file with a dynamic filename based on the variable name
  ggsave(filename = paste0("cluster_proportions_", col, ".png"), 
         plot = categorical_plot, 
         width = 8, height = 6, dpi = 300)
}

```

Education levels are evenly distributed across the clusters. Looking at marital status, the only discrepancy is within the married category, where cluster 4 takes up a significantly higher proportion. Specifically, this means that cluster 3 therefore takes up a significantly smaller proportion. Next, it is noticeable that cluster 1 takes up a small proportion of each of the categories, similar to cluster 4. Regarding income, cluster 2 takes up the biggest proportion of 'less than 40K' category, followed by '40K-60K', and evenly distributed across the rest. Conversely, cluster 3 takes up the largest proportion of the '120K+' category, followed by '80K-120K' and decreasing with decreasing income. Looking at card categories, clusters are evenly distributed across 'Blue', whereas no members of cluster 2 own any other card categories. Furthermore, no members of cluster 1 own 'platinum' cards. Very little members of cluster 1 own 'gold' and 'silver' cards.  

This information further reinforces the earlier built profiles. 

We can also look at the PCA loadings for each of the principal components to better understand the 3D cluster visualisation. 

```{r}

# Get PCA loadings
pca_loadings <- as.data.frame(pca$rotation[, 1:3])  # First two PCs

# Sort by highest absolute contribution
pca_loadings$Feature <- rownames(pca_loadings)
pca_loadings <- pca_loadings[order(abs(pca_loadings$PC1), decreasing = TRUE), ]

# Print most important features for PCs
head(pca_loadings, 10)

pca_var_plot <- fviz_pca_var(pca, col.var = "contrib") +
  theme_minimal() +
  labs(title = "PCA Variable Contributions")

# Display the plot in the R Markdown document
print(pca_var_plot)

# Save the plot as a PNG file
ggsave("pca_variable_contributions.png", plot = pca_var_plot, width = 8, height = 6, dpi = 300)

```

##Visualise means for clustering 

```{r}

# Compute means per cluster
num_summary <- df_clustering %>%
  group_by(Cluster) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  pivot_longer(-Cluster, names_to = "Variable", values_to = "Mean_Value")

#Plot
cluster_profile_plot <- ggplot(num_summary, aes(x = Variable, y = Mean_Value, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Cluster Profiles: Mean of Key Variables", x = "Variable", y = "Mean Value")

# Display the plot in the R Markdown document
print(cluster_profile_plot)

# Save the plot as a PNG file
ggsave("cluster_profiles_mean_values.png", plot = cluster_profile_plot, width = 10, height = 6, dpi = 300)

```
## Hierarchical clustering

It is also useful to look at hierarchical clustering in addition to k-means clustering to compare results. 

```{r}

# Use only numerical features for simplicity 
df_hclust <- df_clustering %>% select_if(is.numeric)

# Compute distance matrix
dist_matrix <- dist(df_clustering, method = "euclidean")

# Perform hierarchical clustering using Ward's method
hc_model <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram for hierarchical clustering model
plot(hc_model, cex = 0.6, hang = -1, main = "Hierarchical Clustering Dendrogram")

# Save the plot as a PNG file
png("hierarchical_clustering_dendrogram.png", width = 800, height = 600)
plot(hc_model, cex = 0.6, hang = -1, main = "Hierarchical Clustering Dendrogram")
dev.off()

# Cut tree into K clusters 
k <- 4
rect.hclust(hc_model, k = k, border = 2:5)

# Add cluster assignments to your dataset
hc_clusters <- cutree(hc_model, k = k)
df_clustering$hc_cluster <- as.factor(hc_clusters)

```
The dendogram shows 4 clear clusters as a result, separated by different colours. Since it is not clear to see the characteristics of these clusters, we move to further visualise them for interpretation. 

### Visualise clusters 

```{r}

# Perform PCA
pca_res <- prcomp(df_hclust, center = TRUE, scale. = TRUE)

# Extract the first three PCs
pca_df <- as.data.frame(pca_res$x[, 1:3])  # Using the first three principal components
pca_df$Cluster <- as.factor(hc_clusters)   # Assuming 'hc_clusters' holds your cluster labels

# Create a 3D plot using plotly
p <- plot_ly(data = pca_df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~Cluster, 
             colors = c("red", "blue", "green", "purple"), 
             type = "scatter3d", mode = "markers", marker = list(size = 5)) %>%
  layout(title = "PCA of Hierarchical Clustering (3D)",
         scene = list(xaxis = list(title = "PC1"),
                      yaxis = list(title = "PC2"),
                      zaxis = list(title = "PC3")))

# Show plot
p

```

```{r}
# PCA for visualization
pca_res <- prcomp(df_hclust)

# Create the hierarchical clustering PCA plot and store it in a variable
hcluster_plot <- fviz_cluster(list(data = df_hclust, cluster = hc_clusters), 
                             geom = "point", ellipse.type = "convex", 
                             palette = "jco", ggtheme = theme_minimal(),
                             main = "Hierarchical Clustering - PCA Plot")

# Display the plot in the R Markdown document
print(hcluster_plot)

# Save the plot as a PNG file
ggsave("hierarchical_clustering_pca_plot.png", plot = hcluster_plot, width = 10, height = 8, dpi = 300)

```


```{r}

# Compute silhouette scores for hierarchical clustering
sil <- silhouette(cutree(hc_model, k = 4), dist_matrix)

# Create the silhouette plot and store it in a variable
hsilhouette_plot <- fviz_silhouette(sil) +
  ggtitle("Silhouette Plot - Hierarchical Clustering")

# Display the silhouette plot in the R Markdown document
print(hsilhouette_plot)

# Save the plot as a PNG file
ggsave("silhouette_plot_hierarchical_clustering.png", plot = hsilhouette_plot, width = 8, height = 6, dpi = 300)

```
The silhouette widths are relatively low, indicating that the clusters may not be well separated. 

### Cluster interpretation

```{r}

df_clustered <- df_clustering %>%
  mutate(hc_cluster = factor(cutree(hc_model, k = 4)))

# Summary by cluster
df_clustered %>%
  group_by(hc_cluster) %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd)))

```

#Linear regression

We use linear regression in this case to predict revolving balance and what factors affect it. 

##Preparation of data

Firstly, we prepare the data for regression by removing irrelevant, highly correlated variables and handling categorical variables to be used for analysis. 

```{r}
# Select predictors (excluding the dependent variable from the dataset)
df_reg <- clean_data %>%
  dplyr::select(-c(CLIENTNUM, Customer_Age, Total_Amt_Chng_Q4_Q1,Total_Revolving_Bal, Avg_Open_To_Buy, Total_Trans_Ct))  # Exclude ID and target variable

# Define the dependent variable
y <- clean_data$Total_Revolving_Bal

# Ensure categorical variables are factors
cat_cols <- names(df_reg)[sapply(df_reg, is.character)]
df_reg[cat_cols] <- lapply(df_reg[cat_cols], as.factor)

# Set proper contrasts for categorical variables (drops first level as reference)
for (col in cat_cols) {
  contrasts(df_reg[[col]]) <- contr.treatment(nlevels(df_reg[[col]]))
}

# Check data structure
str(df_reg)

```

We define the base model and the full model in order to begin with stepwise regression. 

```{r}
X <- df_reg

# Fit full model with all predictors
full_model <- lm(y ~ ., data = df_reg)

# Fit null model (intercept only)
null_model <- lm(y ~ 1, data = df_reg)

summary(full_model)
```

##Forward stepwise 

```{r}
forward_model <- step(null_model, 
                      scope = list(lower = null_model, upper = full_model), 
                      direction = "forward", 
                      trace = FALSE)

summary(forward_model)

```

##Backward stepwise

```{r}
backward_model <- step(full_model, 
                       direction = "backward", 
                       trace = FALSE)

summary(backward_model)

```

##Best stepwise

```{r}
both_model <- step(null_model, 
                   scope = list(lower = null_model, upper = full_model), 
                   direction = "both", 
                   trace = FALSE)

summary(both_model)

```

##Stepwise model comparison

```{r}
# Extract selected variables from each model
forward_vars <- names(coef(forward_model))[-1]  # Exclude intercept
backward_vars <- names(coef(backward_model))[-1]
both_vars <- names(coef(both_model))[-1]

# Create a comparison table
selected_vars <- data.frame(
  Variable = unique(c(forward_vars, backward_vars, both_vars)),
  Forward = ifelse(unique(c(forward_vars, backward_vars, both_vars)) %in% forward_vars, "✅", "❌"),
  Backward = ifelse(unique(c(forward_vars, backward_vars, both_vars)) %in% backward_vars, "✅", "❌"),
  Both = ifelse(unique(c(forward_vars, backward_vars, both_vars)) %in% both_vars, "✅", "❌")
)

print(selected_vars)

```

```{r}
# Compute RMSE for each model
rmse_forward <- sqrt(mean(forward_model$residuals^2))
rmse_backward <- sqrt(mean(backward_model$residuals^2))
rmse_both <- sqrt(mean(both_model$residuals^2))

# Compare AIC and BIC values
model_performance <- data.frame(
  Model = c("Forward Stepwise", "Backward Stepwise", "Bidirectional Stepwise"),
  RMSE = c(rmse_forward, rmse_backward, rmse_both),
  AIC = c(AIC(forward_model), AIC(backward_model), AIC(both_model)),
  BIC = c(BIC(forward_model), BIC(backward_model), BIC(both_model))
)

print(model_performance)

# Create the bar plot for model performance
model_performance_plot <- ggplot(model_performance, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Stepwise Model Performance Comparison", x = "Model", y = "RMSE")

# Display the plot in the R Markdown document
print(model_performance_plot)

# Save the plot as a PNG file
ggsave("stepwise_model_performance_comparison.png", plot = model_performance_plot, width = 10, height = 6, dpi = 300)

```

All models seem to have the same RMSE, AIC and BIC due to selection of the same features. The R squared also seems to be relatively low considering we are using all the variables in the dataset. 

To better understand why the stepwise models are not performing well, we check for VIF values for the variables. 

```{r}

# Compute Variance Inflation Factor (VIF)
vif_values <- vif(full_model)
print(vif_values)

# Filter variables with VIF > 10 (high collinearity)
high_vif_vars <- names(vif_values[vif_values > 10])
print(high_vif_vars)
```
All of the variables seem to have relatively low VIF values, aside from gender and income category, which seem to have higher collinearity with other variables in the dataset. This could be causing issues with the linear models. 

```{r}
# Check for aliased variables
alias(full_model)

```

Since there are some correlations in the model that cannot be accounted for through stepwise models, we proceed with regularisation to deal with multicolinearity in the data. 

##Regularisation methods 
###Prepare data

Firstly, we split the data into training and testing, while also handling categorical variables. 

```{r}
# Convert categorical variables to dummies
df_dummies <- model.matrix(~ . - 1, data = df_reg)  # Creates dummy variables

# Define dependent variable
y <- clean_data$Total_Revolving_Bal
X <- as.matrix(df_dummies)  # Convert to matrix

# Split into training & test sets
set.seed(42)
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

```

###Lasso regression

```{r}
# Fit LASSO model
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1)  # LASSO (alpha=1)

plot(lasso_model)

# Open PNG device to save plot
png("lasso_model_plot.png", width = 800, height = 600)

# Generate and print the plot (this assumes you have already created the plot)
plot(lasso_model)

# Close the PNG device to save the file
dev.off()

best_lambda_lasso <- lasso_model$lambda.min
print(paste("Best Lambda for Lasso:", best_lambda_lasso))

# Predict
lasso_pred <- predict(lasso_model, s = best_lambda_lasso, newx = X_test)

# Model performance
lasso_rmse <- sqrt(mean((lasso_pred - y_test)^2))
print(paste("LASSO RMSE:", lasso_rmse))

#Get coefficients
lasso_coefs <- coef(lasso_model, s=best_lambda_lasso)
print(lasso_coefs)
```

###Ridge regression

```{r}
# Fit Ridge model
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0)  # Ridge (alpha=0)

# Plot Lambda selection
plot(ridge_model) 
best_lambda_ridge <- ridge_model$lambda.min
print(paste("Best Lambda for Ridge:", best_lambda_ridge))

# Open PNG device to save plot
png("ridge_model_plot.png", width = 800, height = 600)

# Generate and print the plot (this assumes you have already created the plot)
plot(ridge_model)

# Close the PNG device to save the file
dev.off()

# Predict
ridge_pred <- predict(ridge_model, s = best_lambda_ridge, newx = X_test)

# Model performance
ridge_rmse <- sqrt(mean((ridge_pred - y_test)^2))
print(paste("Ridge RMSE:", ridge_rmse))

# Get coefficients
ridge_coefs <- coef(ridge_model, s = best_lambda_ridge)
print(ridge_coefs)

```


###Elastic net

```{r}
# Fit Elastic Net model (mix of LASSO & Ridge)
elastic_model <- cv.glmnet(X_train, y_train, alpha = 0.5)  # Elastic Net (alpha=0.5)

# Plot Lambda selection
plot(elastic_model) 
best_lambda_elastic <- elastic_model$lambda.min
print(paste("Best Lambda for Elastic Net:", best_lambda_elastic))

# Open PNG device to save plot
png("elastic_model_plot.png", width = 800, height = 600)

# Generate and print the plot (this assumes you have already created the plot)
plot(elastic_model)

# Close the PNG device to save the file
dev.off() 

# Predict
elastic_pred <- predict(elastic_model, s = best_lambda_elastic, newx = X_test)

# Model performance
elastic_rmse <- sqrt(mean((elastic_pred - y_test)^2))
print(paste("Elastic Net RMSE:", elastic_rmse))

# Get coefficients
elastic_coefs <- coef(elastic_model, s = best_lambda_elastic)
print(elastic_coefs)
```

###Comparison

```{r}
# Compare RMSE of all models
model_performance <- data.frame(
  Model = c("LASSO", "Ridge", "Elastic Net"),
  RMSE = c(
    lasso_rmse,
    ridge_rmse,
    elastic_rmse
  )
)

# Print performance
print(model_performance)

# Create the plot comparing model performance
regularisation_model_performance_plot <- ggplot(model_performance, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Model Performance Comparison", x = "Model", y = "RMSE")

# Print the plot to display in the console or R Markdown
print(regularisation_model_performance_plot)

# Save the plot as a PNG file
ggsave("regularisation_model_performance_comparison.png", plot = regularisation_model_performance_plot, width = 8, height = 6, dpi = 300)

```

###Hybrid approach

We also try a hybrid approach of stepwise feature selection with Ridge regression to then shrink the coefficients. Since Lasso removes many of the categories of the same variable, it is relevant to find an optimal solution between feature selection and regularisation. 

```{r}
# Get Ridge-selected features (keep those with non-negligible coefficients)
ridge_selected <- which(abs(coef(ridge_model, s = best_lambda_ridge)) > 0.01)
ridge_selected_vars <- rownames(coef(ridge_model))[ridge_selected]
print("Selected Variables from Ridge:")
print(ridge_selected_vars)

ridge_selected_vars <- ridge_selected_vars[ridge_selected_vars != "(Intercept)"]

df_ridge_selected <- as.data.frame(df_dummies)
# Subset dataset with only Ridge-selected variables
df_ridge_selected <- df_ridge_selected %>%
  dplyr::select(all_of(ridge_selected_vars))

# Fit stepwise model using Ridge-selected features
stepwise_ridge <- step(lm(y ~ ., data = df_ridge_selected), 
                        direction = "both", trace = FALSE)

summary(stepwise_ridge)

```
The results are similar to the original stepwise models, hence not much improvement has been made to the model with regularisation. 

##Random Forest

Since other models didn't work due to multicolinearity in the model, we turn to random forest to deal with this multicolinearity and do regression using a model that's good at dealing with this - random forest. 

```{r}

# Define predictors (X) and target variable (Y)
X_rf <- clean_data %>% dplyr::select(-c(Total_Revolving_Bal,CLIENTNUM))  # Exclude target variable
y_rf <- clean_data$Total_Revolving_Bal  # Define dependent variable

# Convert categorical variables to factors if necessary
cat_cols <- names(X_rf)[sapply(X_rf, is.character)]
X_rf[cat_cols] <- lapply(X_rf[cat_cols], as.factor)

# Fit Random Forest model
set.seed(123)  # For reproducibility
rf_model <- randomForest(x = X_rf, y = y_rf, 
                         ntree = 500,  # Number of trees
                         mtry = sqrt(ncol(X_rf)),  # Number of variables considered at each split
                         importance = TRUE)  # Track variable importance

# Print model summary
print(rf_model)

```
It is clear that the random forest model has drastically decreased the RMSE as well as increased the R-squared value by more than 20%. It is clear that the model is performing significantly better than either of the previous regression models. 

```{r}

# Define custom AIC and BIC functions for non-likelihood models like Random Forest
compute_AIC <- function(n, mse, k) {
  return(n * log(mse) + 2 * k)
}

compute_BIC <- function(n, mse, k) {
  return(n * log(mse) + log(n) * k)
}
 
# Predictions on training data
rf_pred <- predict(rf_model, newdata = X_rf)

# Compute RMSE
rmse_rf <- sqrt(mean((rf_pred - y_rf)^2))
print(paste("RMSE for Random Forest:", rmse_rf))

# Compute R²
r_squared_rf <- 1 - sum((rf_pred - y_rf)^2) / sum((y_rf - mean(y_rf))^2)
print(paste("R-squared for Random Forest:", r_squared_rf))

# Compute AIC and BIC (using log-likelihood approximation)
n <- nrow(clean_data)
k_rf <- ncol(X_rf)  # Number of predictors
mse_rf <- mean((rf_pred - y_rf)^2)

aic_rf <- compute_AIC(n, mse_rf, k_rf)
bic_rf <- compute_BIC(n, mse_rf, k_rf)

print(paste("AIC for Random Forest:", aic_rf))
print(paste("BIC for Random Forest:", bic_rf))

```

### Visualise importance 

We plot the feature importance to understand which variables affect the target variable the most. 

```{r}
#Get feature importance
importance(rf_model)

# Visualize feature importance using varImpPlot
varImpPlot(rf_model, main = "Feature Importance - Random Forest")

# Save the feature importance plot
# Open PNG device to save the plot
png("feature_importance_rf_model.png", width = 800, height = 600)

# Print the plot (this will save it to the file opened by png())
varImpPlot(rf_model, main = "Feature Importance - Random Forest")

# Close the device to save the file
dev.off()

```
It is clear to see that Average Utilisation Ratio is by far the most important variable in the model, which is also the variable that consistently had significantly low p-values across previous models. In second and third place are Average Open to Buy and Credit Limit which, similar to Average Utilisation Ratio, has been consistently significant in the previous models. 

```{r}
# Get variable importance data
importance_df <- data.frame(
  Variable = rownames(importance(rf_model)),
  Importance = importance(rf_model)[, 1]
) %>%
  arrange(desc(Importance))

# Create the bar plot for the top 10 most important variables
importance_plot <- ggplot(importance_df[1:10, ], aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#2C3E50") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Most Important Variables (Random Forest)", x = "Variable", y = "Importance")

# Print the plot to display it
print(importance_plot)

# Save the plot as a PNG file
ggsave("top_10_important_variables_rf_r.png", plot = importance_plot, width = 8, height = 6, dpi = 300)

```
It is not only relevant to analyse the importance of the features, but also how they affect the target variable directly. Hence, we plot partial dependence plots for the top variables. 

```{r}

# Partial Dependence Plot for top variables
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Trans_Ct",
            main = "Partial Dependence on Total_Trans_Ct", 
            xlab = "Total Transaction Count", ylab = "Predicted Revolving Balance")

# Partial Dependence Plot for top variables 
partialPlot(rf_model, pred.data = X_rf, x.var = "Attrition_Flag",
            main = "Partial Dependence on Attrition_Flag", 
            xlab = "Attrition Flag", ylab = "Predicted Revolving Balance")

# Partial Dependence Plot for top variables 
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Relationship_Count",
            main = "Partial Dependence on Total_Relationship_Count", 
            xlab = "Total Relationship Count", ylab = "Predicted Revolving Balance")

# Partial Dependence Plot for top variables 
partialPlot(rf_model, pred.data = X_rf, x.var = "Avg_Utilization_Ratio",
            main = "Partial Dependence on Avg_Utilization_Ratio", 
            xlab = "Average Utilization Ratio", ylab = "Predicted Revolving Balance")

# Partial Dependence Plot for top variables 
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Trans_Amt",
            main = "Partial Dependence on Total_Trans_Amt", 
            xlab = "Total Transaction Amount", ylab = "Predicted Revolving Balance")

# Generate and save Partial Dependence Plots for the top variables
png("partial_dependence_Total_Trans_Ct.png", width = 800, height = 600)
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Trans_Ct",
            main = "Partial Dependence on Total_Trans_Ct", 
            xlab = "Total Transaction Count", ylab = "Predicted Revolving Balance")
dev.off()

png("partial_dependence_Attrition_Flag.png", width = 800, height = 600)
partialPlot(rf_model, pred.data = X_rf, x.var = "Attrition_Flag",
            main = "Partial Dependence on Attrition_Flag", 
            xlab = "Attrition Flag", ylab = "Predicted Revolving Balance")
dev.off()

png("partial_dependence_Total_Relationship_Count.png", width = 800, height = 600)
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Relationship_Count",
            main = "Partial Dependence on Total_Relationship_Count", 
            xlab = "Total Relationship Count", ylab = "Predicted Revolving Balance")
dev.off()

png("partial_dependence_Avg_Utilization_Ratio.png", width = 800, height = 600)
partialPlot(rf_model, pred.data = X_rf, x.var = "Avg_Utilization_Ratio",
            main = "Partial Dependence on Avg_Utilization_Ratio", 
            xlab = "Average Utilization Ratio", ylab = "Predicted Revolving Balance")
dev.off()

png("partial_dependence_Total_Trans_Amt.png", width = 800, height = 600)
partialPlot(rf_model, pred.data = X_rf, x.var = "Total_Trans_Amt",
            main = "Partial Dependence on Total_Trans_Amt", 
            xlab = "Total Transaction Amount", ylab = "Predicted Revolving Balance")
dev.off()



```

We analyse the effects of each variable:
Total Transaction Count: Initially, with very low transaction counts there is a sharp increase in predicted revolving balance. At some point, the gradient decreases but overall higher transaction count is associated with higher revolving balance. Around the mean value, this relationship becomes reciprocal before plateauing with very high transaction counts. 
Attrition Flag: Customers that still hold accounts with the bank generally have higher revolving balance. 
Total Relationship Count: With lower than average products owned by the customer, there is a positive relationship between revolving balance and the total relationship count. Around the mean the line becomes flat, meaning that there is no further increase in revolving balance with more products owned, but generally high revolving balance associated with these customers. 
Average Utilisation Ratio: Similar analysis as above. Sharp increase with low levels of utilisation and a generally positive relationship between revolving balance and average utilisation ratio. At some point, the gradient of the line significantly decreases, but overall higher utilisation means higher revolving balance. 
Total Transaction Amount: Similar analysis as above. Low levels of transaction amount mean lower revolving balance. At some 'breaking point' there is a sharp increase in the revolving balance with a small increase in total transaction amount. After this point, the line flattens. Generally, higher transaction amounts mean higher revolving balance of a customer. 

###Residual analysis 

Finally, it is important to analyse the residuals under the assumption that they should be normally distributed with mean 0. 

```{r}
rf_residuals <- y_rf - rf_pred

# Histogram of residuals
ggplot(data.frame(residuals = rf_residuals), aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "#0072B2", color = "white") +
  theme_minimal() +
  labs(title = "Residual Distribution - Random Forest", x = "Residual", y = "Count")

# Open the PNG device and save the histogram
png("residual_histogram_random_forest.png", width = 800, height = 600)

# Generate the histogram plot
ggplot(data.frame(residuals = rf_residuals), aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "#0072B2", color = "white") +
  theme_minimal() +
  labs(title = "Residual Distribution - Random Forest", x = "Residual", y = "Count")

# Close the device to save the plot
dev.off()

```

Residuals are slightly right skewed meaning that it's underestimating the values slightly, although the overall distribution is normal around 0, so a good predictor. 

# Classification

Predicting whether attrition of customers is important for banks to manage credit limits and loans. Thus we perform the classification taks with 'Attrition_Flag' as our target variable. Firstly, since the ratio of Attrited Customers to Existing Customers in the dataset is low and this can impact the accuracy of prediction, we oversample using the ROSE library. This ensures that the ratio is roughly 1:1 in the training dataset. We retain the test data as is, in order to be able to generalise results.  

```{r}

# Split data into training and testing sets
set.seed(123)
trainIndex <- createDataPartition(clean_data$Attrition_Flag, p = 0.8, list = FALSE, times = 1)
train_data <- clean_data[trainIndex, ]
test_data <- clean_data[-trainIndex, ]

# Check the distribution of the target variable in the training set
table(train_data$Attrition_Flag)

# Remove the ClientNum column 
train_data <- train_data %>%
  dplyr::select(-CLIENTNUM)

# Remove the ClientNum column 
test_data <- test_data %>%
  dplyr::select(-CLIENTNUM)

train_data$Gender <- as.factor(train_data$Gender)
train_data$Card_Category <- as.factor(train_data$Card_Category)

test_data$Gender <- as.factor(test_data$Gender)
test_data$Card_Category <- as.factor(test_data$Card_Category)

# Check if 'Attrition_Flag' is a factor (ROSE expects a factor for categorical target)
train_data$Attrition_Flag <- as.factor(train_data$Attrition_Flag)
test_data$Attrition_Flag <- as.factor(test_data$Attrition_Flag)

# Apply oversampling using ROSE to balance the target variable in the training set
train_data_balanced <- ROSE(Attrition_Flag ~ ., data = train_data, seed = 123)$data

# Check the distribution of the target variable in the balanced training set
table(train_data_balanced$Attrition_Flag)
table(test_data$Attrition_Flag)

```
We see that initially, the ratio in the training data is 1:5, and after applying the oversampling it turns to 1:1. For reference, in the test dataset, the ratio is around 1:6. 

## Random Forest Classification

Firstly, we apply the random forest classification as this model performed best on the regression task. 

```{r}
# Ensure consistent factor levels for 'Card_Category' across both training and testing data
all_levels <- union(levels(train_data_balanced$Card_Category), levels(test_data$Card_Category))

# Apply the same levels to both datasets
train_data_balanced$Card_Category <- factor(train_data_balanced$Card_Category, levels = all_levels)
test_data$Card_Category <- factor(test_data$Card_Category, levels = all_levels)

# Fit Random Forest Model
rf_class <- randomForest(Attrition_Flag ~ ., data = train_data_balanced, importance = TRUE)

# Evaluate model
rf_class_pred <- predict(rf_class, newdata = test_data)

# Confusion matrix
rf_confusion <- confusionMatrix(rf_class_pred, test_data$Attrition_Flag)

print(rf_confusion)

varImpPlot(rf_class)

# Open the PNG device to save the plot
png("rf_feature_importance.png", width = 800, height = 600)

# Generate and save the feature importance plot
varImpPlot(rf_class, main = "Feature Importance - Random Forest")

# Close the device to save the plot
dev.off()


```
We see that accuracy, sensitivity and specificity of the model are very high, and the confusion matrix shows that we have less false positives than false negatives. This would be concerning in the context of the issue, since it's better to predict that a customer attrites and they don't, than the other way around. Nevertheless, the performance of the model is very good. 

Looking at the variable importance, total transaction count and amount are both the most important variables in predicting attrition. Total relationship count follows with total amount change from q4 to q1. Other variables important to the model are average utilisation ratio and customer age. All of these make sense in the real world context, where how often the customer utilises the credit card and how many and how big of transaction they make likely makes the customer more or less likely to churn. Similarly, depending on how many products they hold, they are more or less likely to close the account and stop the relationship with the bank. Finally, older customers likely close their account at some point and age seems to be one of the determining factors in the model.

## Logistic Regression 

We move to a logistic regression model, which fits the data to a Sigmoid curve, and we define the 'cutoff' point as 0.5. This means that all the predictions with probability >0.5 will be assigned to Attrited Customer. 

```{r}
# Fit Logistic Regression
logistic_model <- glm(Attrition_Flag ~ ., data = train_data, family = "binomial")

# Evaluate model
logistic_pred <- predict(logistic_model, newdata = test_data, type = "response")
logistic_pred_class <- ifelse(logistic_pred > 0.5, "Attrited Customer", "Existing Customer")

# Confusion matrix
logistic_confusion <- confusionMatrix(factor(logistic_pred_class), test_data$Attrition_Flag)

print(logistic_confusion)

```
This model seems to perform poorly, especially compared to the earlier random forest model. 

## Support Vector Machine 

Finally, we test the support vector machine model to the classificaiton task. Here we train the model using different kernels - radial, linear, polynomial and sigmoid. The best performing in terms of accuracy, sensitivity and specificity is radial. 

```{r}

# Fit SVM model
svm_model <- svm(Attrition_Flag ~ ., data = train_data, type = "C-classification", kernel = "radial")

# Predict
svm_pred <- predict(svm_model, newdata = test_data)

# Confusion matrix
svm_confusion <- confusionMatrix(svm_pred, test_data$Attrition_Flag)

print(svm_confusion)

```
This model also seems to be performing well in terms of accuracy and specificity, although not too high in sensitivity. This is relevant as mentioned earlier, since it is better to have false positives in this scenario than false negatives. 

## Model comparison 

```{r}
# Compare Accuracy, Precision, Recall, and F1 Score for each model
rf_results <- confusionMatrix(rf_class_pred, test_data$Attrition_Flag)
logistic_results <- confusionMatrix(factor(logistic_pred_class), test_data$Attrition_Flag)
svm_results <- confusionMatrix(svm_pred, test_data$Attrition_Flag)

# Summarize
list(RF = rf_results$overall, Logistic = logistic_results$overall, SVM = svm_results$overall)

```

## Precision-recall and ROC curves

We plot the ROC curves for the three models to assess the area under the curve (AUC).

```{r}

# ROC for SVM
roc_svm <- roc(test_data$Attrition_Flag, as.numeric(svm_pred))
auc_svm <- auc(roc_svm)
plot(roc_svm, col = "blue", main = "ROC Curves for SVM, Random Forest and Logistic Regression")

# ROC for Random Forest
roc_rf <- roc(test_data$Attrition_Flag, as.numeric(rf_class_pred))
auc_rf <- auc(roc_rf)
plot(roc_rf, col = "red", add = TRUE)

# ROC for Logistic Regression
roc_lr <- roc(test_data$Attrition_Flag, as.numeric(logistic_pred))
auc_lr <- auc(roc_lr)
plot(roc_lr, col = "purple", add = TRUE)

# Add legend
legend("bottomright", legend = c(paste("SVM (AUC = ", round(auc_svm, 2), ")", sep = ""),
                  paste("Random Forest (AUC = ", round(auc_rf, 2), ")", sep = ""),
                  paste("Logistic Regression (AUC = ", round(auc_lr, 2), ")", sep = "")),
       fill = c("blue", "red","purple"))

# Save the plot to a file
ggsave("ROC_Curve_Comparison.png")

```
The logistic regression had low specificity and sensitivity but has the highest AUC. It is followed by random forest, which not only has good accuracy but also an AUC of 0.90. Support vector machine still has a high AUC, but overall the random forest model clearly performs best. 

```{r}

# Precision-Recall for SVM
pr_svm <- pr.curve(scores.class0 = as.numeric(svm_pred), weights.class0 = test_data$Attrition_Flag == "Attrited Customer", curve = TRUE)

# Precision-Recall for Random Forest
pr_rf <- pr.curve(scores.class0 = as.numeric(rf_class_pred), weights.class0 = test_data$Attrition_Flag == "Attrited Customer", curve = TRUE)

# Precision-Recall for Logistic Regression
pr_lr <- pr.curve(scores.class0 = as.numeric(logistic_pred), weights.class0 = test_data$Attrition_Flag == "Attrited Customer", curve = TRUE)

# Plot all Precision-Recall curves in one plot
plot(pr_svm, main = "Precision-Recall Curves for SVM, Random Forest and Logistic Regression", col = "blue")
plot(pr_rf, add = TRUE, col = "red")
plot(pr_lr, add = TRUE, col = "purple")

# Add a legend
legend("topright", legend = c("SVM", "Random Forest", "Logistic Regression"), fill = c("blue", "red", "purple"))

```
The precision recall curve for the three models are fairly similar. 

```{r}

# Create a summary table with the metrics
model_comparison <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "SVM"),
  Accuracy = c(logistic_confusion$overall['Accuracy'], rf_confusion$overall['Accuracy'], svm_confusion$overall['Accuracy']),
  Precision = c(logistic_confusion$byClass['Pos Pred Value'], rf_confusion$byClass['Pos Pred Value'], svm_confusion$byClass['Pos Pred Value']),
  Recall = c(logistic_confusion$byClass['Sensitivity'], rf_confusion$byClass['Sensitivity'], svm_confusion$byClass['Sensitivity']),
  AUC = c(auc_lr, auc_rf, auc_svm)
)

# Print the table
print(model_comparison)
``` 

```{r}

# Get confusion matrix data
conf_matrix <- rf_confusion$table

# Reshape the confusion matrix to long format for ggplot
conf_matrix_melted <- melt(conf_matrix)
colnames(conf_matrix_melted) <- c("Prediction", "Reference", "Freq")

# Plot confusion matrix as heatmap
ggplot(data = conf_matrix_melted, aes(x = Prediction, y = Reference, fill = Freq)) + 
  geom_tile(color = "white", size = 1) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  geom_text(aes(label = Freq), vjust = 1, color = "black", size = 6) +
  labs(title = "Confusion Matrix for Random Forest", x = "Predicted", y = "Actual") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("Attrited Customer", "Existing Customer")) +
  scale_y_discrete(labels = c("Attrited Customer", "Existing Customer"))

# Save the plot to a file
ggsave("Random_Forest_Confusion_Matrix.png")

```

## Hyperparameter tuning

We try hyperparameter tuning to improve the support vector machine and random forest models. We begin by performing cross validation and tuning the values in the svm model to find the best one. 

```{r}
# Hyperparameter tuning for SVM
train_control <- trainControl(method = "cv", number = 5)

svm_tune <- train(Attrition_Flag ~ ., data = train_data_balanced, method = "svmRadial", trControl = train_control, tuneLength = 5)
print(svm_tune)

```

We perform a similar procedure to the random forest model. 

```{r}

# Random Forest hyperparameter tuning
rf_tune <- train(Attrition_Flag ~ ., data = train_data_balanced, method = "rf", trControl = train_control, tuneLength = 5)

# Best tuning parameters
print(rf_tune)
```

As a result, we see that the accuracy of both model decreases as a result, likely due to overfitting. We conclude that the original random forest model was best in terms of performance and resume with interpreting its results. 

### Interpretation

Finally, we look at how the changes in the most important features affect the probability of attrition. 


```{r}

pdp_plot <- partial(rf_class, pred.var = "Total_Trans_Ct", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Total Transaction Count")

pdp_plot <- partial(rf_class, pred.var = "Total_Revolving_Bal", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Total Revolving Balance")

pdp_plot <- partial(rf_class, pred.var = "Total_Relationship_Count", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Total Relationship Count")

pdp_plot <- partial(rf_class, pred.var = "Total_Trans_Amt", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Total Transaction Amount")

pdp_plot <- partial(rf_class, pred.var = "Avg_Utilization_Ratio", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Average Utilisation Ratio")

pdp_plot <- partial(rf_class, pred.var = "Customer_Age", train = train_data_balanced, prob = TRUE)
plotPartial(pdp_plot, main = "PDP with Probabilities for Customer Age")


```
Total Transaction Count: Curve closely resembles a Sigmoid curve. With low transaction count, customers are more likely to remain with the bank. Around the mean value, this probability sharply increases to almost 80% and then plateaus with high values of transaction count. In other words, customers with a high transaction count are more likely to close the account.
Total Revolving Balance: With low total revolving balance, customers are likely to stay an existing customer. At some point above average total revolving balance significantly increases the likelihood of attrition. With extremely high values, the probability decreases down to roughly 50%, which means they are equally likely to stay and leave. 
Total Relationship Count: Exhibits a strong positive relationship with probability of attrition. Customers that hold less products of the bank are less likely to leave and vice versa. However, the probabilities in the graph only range between 0.3 to 0.6 hence the effect of the total relationship count is not as strong. 
Total Transaction Amount: Similar analysis as total revolving balance. With low values of total transaction amount, customers are less likely to close their account, and this probability increases with an increase in total transaction amount. However, after a certain point, the probability rapidly decreases for large values of transaction amount. 
Average Utilisation Ratio: The graph also resembles a Sigmoid curve, albeit not that closely and in a shorter range. Low utilisation ratio makes custoemrs less likely to close their account. With an increase in average utilisation ratio, there is a sharp increase in probability of attrition. Finally, this probability flattens around 50% for very large values of utilisation.
Customer Age: The curve resembles a negative quadratic curve with a very limited range - 0.03. Initially, probability of attrition is very high and decreases rapidly with an increase in customer age. Around the mean, the probability starts growing rapidly again with an increase in customer age. This means that both very young and very old customers are most likely to close thier accounts. 

```{r}
# List of variables to generate PDPs for
variables <- c("Total_Trans_Ct", "Total_Revolving_Bal", "Total_Relationship_Count",
               "Total_Trans_Amt", "Avg_Utilization_Ratio", "Customer_Age")

# Loop through each variable to generate, save, and print PDPs
for (var in variables) {
  # Compute Partial Dependence Plot
  pdp_plot <- partial(rf_class, pred.var = var, train = train_data_balanced, prob = TRUE)

  # Save the plot as PNG
  png(paste0("PDP_", var, ".png"), width = 800, height = 600)
  plotPartial(pdp_plot, main = paste("PDP with Probabilities for", var))
  dev.off()  # Close the PNG device

  # Print the plot to the screen
  plotPartial(pdp_plot, main = paste("PDP with Probabilities for", var))
}
```

